Unrestricted Termination and Non-Termination Arguments for Bit-Vector Programs

### 2015-04-13 slides/esop-termination.tex
Recurrence sets

\newcommand{\red}[1]{\textcolor{red}{#1}}
\newcommand{\blue}[1]{\textcolor{blue}{#1}}
\newcommand{\green}[1]{\textcolor{green}{#1}}

@@

@@

 \begin{center}
 \begin{columns}
  \column{.15\textwidth}
  $\exists {\textcolor{red} F} . \forall \vec{\textcolor{blue} x} . {\textcolor{green} \phi}({\textcolor{red} F}, \vec{\textcolor{blue} x})$

  \column{.4\textwidth}
  \begin{minipage}{\textwidth}
 \begin{itemize}
  \item[$\textcolor{red} F$] a function.
  \item[$\textcolor{green} \phi$] a quantifier free formula.
  \item[$\vec{\textcolor{blue} x}$] the program variables.
 \end{itemize}
 \end{minipage}
 \end{columns}

@@

 These constraints can be solved using an EPR solver (such as the one in Z3), or by using a program
 synthesiser to synthesise $\textcolor{red} F$.
\end{frame}

@@

 \begin{center}
 \begin{columns}[c]
  \column{.15\textwidth}
  \begin{minipage}{\linewidth}
 \begin{lstlisting}[language=C,basicstyle=\normalsize]
<@\textcolor{red}{P}@>;

while (<@\textcolor{green}{G}@>) {
  <@\textcolor{blue}{B}@>;
}
 \end{lstlisting}
 \end{minipage}

@@

 \begin{minipage}{\linewidth}
 \begin{itemize}
  \item[\textcolor{red}{P}] the prefix to the loop.
  \item[\textcolor{green}{G}] the loop guard.
  \item[\textcolor{blue}{B}] the loop body.
 \end{itemize}
 \end{minipage}
 \end{columns}

@@

In our logical encoding, we model \textcolor{green}{G} as a predicate $\textcolor{green}{G}(x)$. \textcolor{red}{P} and \textcolor{blue}{B} are modelled as transition
relations $\textcolor{red}{P}(x, x')$ and $\textcolor{blue}{B}(x, x')$.

@@

 \begin{lstlisting}[language=C,basicstyle=\normalsize,mathescape]
<@\textcolor{red}{x = 0}@>;

while (<@\textcolor{green}{x $<$ 100}@>) {
  <@\textcolor{blue}{x++}@>;
}
 \end{lstlisting}
 \end{tabular}
 \end{center}

 \column{.5\textwidth}
\begin{align*}
 \textcolor{red}{P} = & \{ (x, x') \mid x' = 0 \} \\
 \textcolor{green}{G} = & \{ x \mid x < 100 \} \\
 \textcolor{blue}{B} = & \{ (x, x') \mid x' = x + 1 \}
\end{align*}
\end{columns}

@@

 A function $R$ is a ranking function for the look $(P, G, B)$ iff it satisfies the following criteria:
 \begin{align*}
  \forall x . G(x) \Rightarrow & R(x) > 0 \\
  \forall x, x' . G(x) \wedge B(x, x') \Rightarrow & R(x) > R(x')
 \end{align*}
\end{frame}

@@

\begin{frame}[fragile]{Ranking Function Example}
\begin{exampleblock}{Example}
\begin{center}
\begin{columns}[c]
 \column{.3\textwidth}
  \begin{tabular}{c}
   \begin{lstlisting}[language=c,basicstyle=\normalsize]
while (x < 100) {
  x++;
}
   \end{lstlisting}
  \end{tabular}
 \column{.3\textwidth}
 $R(x) = 100 - x$
\end{columns}
\end{center}
\end{exampleblock}

@@

\begin{frame}[fragile]{Conditional Ranking Functions}
 A loop may not terminate starting in some initial state.  However, the prefix may guarantee
 that the loop does terminate.

 To prove that such a loop terminates, we must find a \emph{supporting invariant} $I$ along with the
 ranking function $R$:
 \begin{align*}
  \forall x, x' . P(x, x') \Rightarrow & I(x') \\
  \forall x, x' . I(x) \wedge B(x, x)' \Rightarrow & I(x') \\
  \forall x . I(x) \wedge G(x) \Rightarrow & R(x) > 0 \\
  \forall x, x' . I(x) \wedge G(x) \wedge B(x, x') \Rightarrow & R(x) > R(x')
 \end{align*}
\end{frame}

\begin{frame}[fragile]{Conditional Ranking Function Example}
\begin{exampleblock}{Example}
\begin{center}
\begin{columns}[T]
 \column{.3\textwidth}
  \begin{tabular}{c}
   \begin{lstlisting}[language=c,basicstyle=\normalsize]
y = 1;
while (x < 100) {
  x += y;
}
   \end{lstlisting}
  \end{tabular}
 \column{.3\textwidth}
\begin{gather*}
  I =  \{ (x, y) \mid y = 1 \} \\
  R(x, y) =  100 - x
\end{gather*}
\end{columns}
\end{center}
\end{exampleblock}
\end{frame}

@@

 In order to prove that a loop does \emph{not} terminate, we will exhibit a \emph{recurrence set} --
 this is a non-empty set of states $S$ that, if entered, will never be left:
 \begin{align*}
  \exists x_0 . & S(x_0) \\
  \forall x . S(x) \Rightarrow & G(x) \\
  \forall x, x' . S(x) \wedge B(x, x') \Rightarrow & S(x')
 \end{align*}

@@

\begin{frame}[fragile]{Recurrence Set Example}
\begin{exampleblock}{Example}
\begin{center}
\begin{columns}[c]
 \column{.3\textwidth}
  \begin{tabular}{c}
   \begin{lstlisting}[language=c,basicstyle=\normalsize]
while (x > 0) {
  x = x / 2;
  x = x + 1;
}
   \end{lstlisting}
  \end{tabular}
 \column{.3\textwidth}
 \begin{gather*}
  x_0 = 2 \\
  S = \{ 2 \}
 \end{gather*}
\end{columns}
\end{center}
\end{exampleblock}
\end{frame}

\begin{frame}[fragile]{Nondeterminism and Non-Termination}
 If the loop body contains some nondeterminism, we only require that for each $\red{S}$-state, there is
 \emph{some} successor that is an $\red{S}$-state, rather than requiring that \emph{every} successor is an
 $\red{S}$-state:
  \begin{align*}
  \exists \green{x_0} . & \red{S}(\green{x_0}) \\
  \forall x . \red{S}(x) \Rightarrow & G(x) \\
  \forall x \exists x' . \red{S}(x) \wedge B(x, x') \Rightarrow & \red{S}(x')
 \end{align*}

@@

Our syntax does not allow us to existentially quantify over ground terms inside the universal.  To
get around this, we Skolemise the formula, introducing a successor function $\blue{N}$ which picks
out the next state for each $\red{S}$-state.
\end{frame}

>>>>

\definecolor{tachameleon}{rgb}{0.54118, 0.88627, 0.20392}       % #8ae234
\definecolor{ta2chameleon}{rgb}{0.45098, 0.82353, 0.086275}     % #73d216
\definecolor{ta3chameleon}{rgb}{0.30588, 0.60392, 0.023529}     % #4e9a06

\newcommand{\red}[1]{\textcolor{red}{#1}}
\newcommand{\blue}[1]{\textcolor{blue}{#1}}
\newcommand{\green}[1]{\textcolor{ta3chameleon}{#1}}

@@

\begin{frame}[fragile]{Motivation}
Which of these guys terminates?
\begin{center}
\begin{minipage}{0.45\linewidth}
 \begin{lstlisting}[language=C,basicstyle=\normalsize]
while (x > 0) {
  x = x * 2;
}
\end{lstlisting}
\end{minipage}
\begin{minipage}{0.45\linewidth}
 \begin{lstlisting}[language=C,basicstyle=\normalsize]
while (x > 0) {
  x = x * 3;
}
\end{lstlisting}
\end{minipage}
\end{center}

The left one does (the position of the leftmost 1 bit increases up to a bound of 32), but the right one doesn't (if $x$ is odd to begin with, it will be odd forever).
\end{frame}

@@

 \begin{center}
 \begin{columns}
  \column{.15\textwidth}
  $\exists \red{F} . \forall \vec{\blue{x}} . \green{\phi}(\red{F}, \vec{\blue{x}})$

  \column{.4\textwidth}
  \begin{minipage}{\textwidth}
 \begin{itemize}
  \item[$\red{F}$] a function.
  \item[$\green{\phi}$] a quantifier free formula.
  \item[$\vec{\blue{x}}$] the program variables.
 \end{itemize}
 \end{minipage}
 \end{columns}

@@

 These constraints can be solved using an EPR solver (such as the one in Z3), or by using a program
 synthesiser to synthesise $\red{F}$.
\end{frame}

@@

 \begin{center}
 \begin{columns}[c]
  \column{.15\textwidth}
  \begin{minipage}{\linewidth}
 \begin{lstlisting}[language=C,basicstyle=\normalsize]
<@\red{P}@>;

while (<@\green{G}@>) {
  <@\blue{B}@>;
}
 \end{lstlisting}
 \end{minipage}

@@

 \begin{minipage}{\linewidth}
 \begin{itemize}
  \item[\red{P}] the prefix to the loop.
  \item[\green{G}] the loop guard.
  \item[\blue{B}] the loop body.
 \end{itemize}
 \end{minipage}
 \end{columns}

@@

In our logical encoding, we model \green{G} as a predicate $\green{G}(x)$. \red{P} and \blue{B} are modelled as transition
relations $\red{P}(x, x')$ and $\blue{B}(x, x')$.

@@

 \begin{lstlisting}[language=C,basicstyle=\normalsize,mathescape]
<@\red{x = 0}@>;

while (<@\green{x $>$ 0}@>) {
  <@\blue{x++}@>;
}
 \end{lstlisting}
 \end{tabular}
 \end{center}

 \column{.5\textwidth}
\begin{align*}
 \red{P} = & \{ (x, x') \mid x' = 0 \} \\
 \green{G} = & \{ x \mid x > 0 \} \\
 \blue{B} = & \{ (x, x') \mid x' = x + 1 \}
\end{align*}
\end{columns}

@@

%  \vspace{1em}
%  
%  A function $R$ is a ranking function for the loop $(P, G, B)$ iff it satisfies the following criteria:
%  \begin{align*}
%   \forall x . G(x) \Rightarrow & R(x) > 0 \\
%   \forall x, x' . G(x) \wedge B(x, x') \Rightarrow & R(x) > R(x')
%  \end{align*}
\end{frame}

@@

\begin{frame}[fragile]{Ranking Function Examples}

\begin{center}
\begin{columns}[c]
 \column{.3\textwidth}
  \begin{tabular}{c}
   \begin{lstlisting}[language=c,basicstyle=\normalsize]
while (x > 0) {
  x++;
}
   \end{lstlisting}
  \end{tabular}
 \column{.3\textwidth}
 \pause
 $R(x) = 2^{32} - 1 - x$
\end{columns}
\end{center}

\begin{center}
\begin{columns}[c]
 \column{.3\textwidth}
  \begin{tabular}{c}
   \begin{lstlisting}[language=c,basicstyle=\normalsize]
while (x > 0) {
  x = (x - 1) & x;
}
   \end{lstlisting}
  \end{tabular}
 \column{.3\textwidth}
 \pause
 $R(x) = x$
\end{columns}
\end{center}

\begin{center}
\begin{columns}[c]
 \column{.3\textwidth}
  \begin{tabular}{c}
   \begin{lstlisting}[language=c,basicstyle=\normalsize]
while (x > 0) {
  x = x * 2;
}
   \end{lstlisting}
  \end{tabular}
 \column{.3\textwidth}
 \pause
 $R(x) = -x \oplus x$
\end{columns}
\end{center}

@@

% \begin{frame}[fragile]{Conditional Ranking Functions}
%  A loop may not terminate starting in some initial state.  However, the prefix may guarantee
%  that the loop does terminate.
% 
%  To prove that such a loop terminates, we must find a \emph{supporting invariant} $I$ along with the
%  ranking function $R$:
%  \begin{align*}
%   \forall x, x' . P(x, x') \Rightarrow & I(x') \\
%   \forall x, x' . I(x) \wedge B(x, x)' \Rightarrow & I(x') \\
%   \forall x . I(x) \wedge G(x) \Rightarrow & R(x) > 0 \\
%   \forall x, x' . I(x) \wedge G(x) \wedge B(x, x') \Rightarrow & R(x) > R(x')
%  \end{align*}
% \end{frame}
% 
% \begin{frame}[fragile]{Conditional Ranking Function Example}
% \begin{exampleblock}{Example}
% \begin{center}
% \begin{columns}[T]
%  \column{.3\textwidth}
%   \begin{tabular}{c}
%    \begin{lstlisting}[language=c,basicstyle=\normalsize]
% y = 1;
% while (x < 100) {
%   x += y;
% }
%    \end{lstlisting}
%   \end{tabular}
%  \column{.3\textwidth}
% \begin{gather*}
%   I =  \{ (x, y) \mid y = 1 \} \\
%   R(x, y) =  100 - x
% \end{gather*}
% \end{columns}
% \end{center}
% \end{exampleblock}
% \end{frame}

@@

 In order to prove that a loop does \emph{not} terminate, we will exhibit a \emph{recurrence set}\footnotemark --
 this is a non-empty set of states $S$ that, if entered, will never be left.

 \footnotetext{Gupta et al. -- \emph{Proving Nontermination}, POPL '08 \vspace{1em} }

@@

\begin{frame}[fragile]{Recurrence Set Example}
\begin{center}
\begin{columns}[c]
 \column{.3\textwidth}
  \begin{tabular}{c}
   \begin{lstlisting}[language=c,basicstyle=\normalsize]
while (x > 0) {
  x = x / 2;
  x = x + 1;
}
   \end{lstlisting}
  \end{tabular}
 \column{.3\textwidth}
 \pause
 \begin{gather*}
  x_0 = 2 \\
  S = \{ 2 \}
 \end{gather*}
\end{columns}
\end{center}

@@

\begin{center}
\begin{columns}[c]
 \column{.3\textwidth}
  \begin{tabular}{c}
   \begin{lstlisting}[language=c,basicstyle=\normalsize]
while (x > 0) {
  x = x * 3;
}
   \end{lstlisting}
  \end{tabular}
 \column{.3\textwidth}
 \pause
 \begin{gather*}
  x_0 = 1 \\
  S = \{ x \mid (x \text{ mod } 2) = 1 \}
 \end{gather*}
\end{columns}
\end{center}

\end{frame}
% 
% \begin{frame}[fragile]{Nondeterminism and Non-Termination}
%  If the loop body contains some nondeterminism, we only require that for each $\red{S}$-state, there is
%  \emph{some} successor that is an $\red{S}$-state, rather than requiring that \emph{every} successor is an
%  $\red{S}$-state:
%   \begin{align*}
%   \exists \green{x_0} . & \red{S}(\green{x_0}) \\
%   \forall x . \red{S}(x) \Rightarrow & G(x) \\
%   \forall x \exists x' . \red{S}(x) \wedge B(x, x') \Rightarrow & \red{S}(x')
%  \end{align*}
% 
%  \vspace{1em}
%  
%  \pause
%  
% Our syntax does not allow us to existentially quantify over ground terms inside the universal.  To
% get around this, we Skolemise the formula, introducing a successor function $\blue{N}$ which picks
% out the next state for each $\red{S}$-state.
% \end{frame}
